### 2021-07-20 paper/localpsf.tex
wrote rational positive definite modification section

In order to use the product-convolution approximation $\widetilde{\mathcal{A}}$, we convert $\widetilde{\mathbf{A}}$ (the discretized version of $\widetilde{\mathcal{A}}$) to hierarchical matrix (H-matrix) format. Many classes of hierarchical matrices exist---H1, uniform H1, H2, HODLR, HSS, HBS, and more \cite{HMATRIX}. We use H1 matrices in our numerical results, but any of the other H-matrix formats could be used instead. The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(N^2)$ or $O(N^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(N \log^a N)$ time and memory for hierarchical matrices. Here $N$ is the number of degrees of freedom for the discretized version of the problem, and $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix.
\end{itemize}
For more details on H-matrices, we recommend \cite{HMATRIXGOOD}.

@@

>>>>

We convert $\widetilde{\mathbf{A}}$ (the discretized version of $\widetilde{\mathcal{A}}$) to hierarchical matrix (H-matrix) format. Once in H-matrix format, useful matrix operations such as matrix factorization and inversion may be performed quickly using fast H-matrix methods. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead. The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(N^2)$ or $O(N^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(N \log^a N)$ time and memory for hierarchical matrices. Here $N$ is the number of degrees of freedom for the discretized version of the problem, and $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix.
\end{itemize}
For more details on H-matrices, we recommend \cite{HMATRIXGOOD}.

@@

\subsection{Rational positive semi-definite modification}

In many problems of practical interest (e.g., Hessian approximation) $\mathcal{A}$ is symmetric positive semi-definite. However, $\widetilde{\mathbf{A}}$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable because symmetry and positive semi-definiteness are important properties of an operator which should be preserved by approximations, if possible. Moreover, a lack of symmetry and/or positive semi-definiteness in $\widetilde{\mathbf{A}}$ may prevent one from using many popular and highly effective algorithms to perform further useful matrix operations involving $\widetilde{\mathbf{A}}$.
To address this issue, here we describe a method for modifying $\widetilde{\mathbf{A}}$ to force it to be symmetric positive semi-definite.

We use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the spectrum of $\widetilde{\mathbf{A}}_\text{sym}^+$ has certain desirable properties, which we will describe later in this section. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing the H-matrix inverse $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should typically be avoided, here it is fine to compute the inverse because we are inverting an H-matrix, and there are fast, stable, scalable, and widely accepted algorithms for inverting H-matrices \cite{Hmatrixinverse}.

Our choice of the constants $c_0$, $c_1$, $c_2$, and $\mu$ will depend on upper and lower bounds for the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$. Let $m$ and $M$ be the smallest and largest eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, respectively, so that the interval $[m, M] \subset \mathbb{R}$ contains the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$. Note that $m < 0$ since $\widetilde{\mathbf{A}}_\text{sym}$ is indefinite. 
%Also note that $|m| \ll |M|$ if $\widetilde{A}$ is positive definite, since negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ are a result of approximation error. 
We estimate $m$ and $M$ using the implicitly restarted Lanczos method [CITE], a matrix-free Krylov method. This is fast because computing matrix-vector products with an H-matrix is fast, and because the eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ do not cluster near $m$ or $M$ (though the eigenvalues may cluster at zero). Other matrix-free extremal eigenvalue methods, such as LOBPCG or shifted power iteration, could also be used.

The rational matrix approximation \eqref{eq:rational_matrix_function} may be written as $\widetilde{\mathbf{A}}_\text{sym}^+ = f(\widetilde{\mathbf{A}}_\text{sym})$, where $f$ is the rational function 
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose the four unknown constants, $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of four equations:
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(m) &= |m| \label{eq:desired_property_3} \\
f(M) &= M \label{eq:desired_property_4}
\end{align}
with the further condition that $\mu > |m|$ (so that the pole of $f$ is not in $[m,M]$). Together, these equations imply that $f(0)=0$ is the unique minimum of $f$ on $[m,M]$, which implies that $f$ is non-negative on $[m,M]$, which implies that  $f(\widetilde{\mathbf{A}}_\text{sym}) = \widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. The equation $f(0)=0$ implies that the null space of $\widetilde{\mathbf{A}}_\text{sym}^+$ contains the null-space of $\widetilde{\mathbf{A}}_\text{sym}$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. The equation $f(M)=M$ forces the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ to be close to the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$. I.e., $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. The equation $f(m)=|m|$ ensures that the negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(m)$, then the positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ will be closer to the corresponding positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ (good), but the erroneous negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ will be amplified more in magnitude (bad). We find that $f(m)=|m|$ is an appropriate balance between these competing interests.

Using the definition of $f$ in \eqref{eq:rational_function_scalar}, and basic calculus and linear algebra, we may write Equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, and \eqref{eq:desired_property_4} in the following matrix form:
\begin{equation}
\label{eq:rational_3_linear_system}
\begin{bmatrix} 1 & 0 & \frac{1}{\mu} \\ 0 & 1 & -\frac{1}{\mu^2} \\ 1 & M & \frac{1}{M+\mu} \end{bmatrix} \begin{bmatrix} c_0 \\ c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \\ M \end{bmatrix},
\end{equation}
and we may write Equation \eqref{eq:desired_property_3} as follows: 
\begin{equation}
\label{eq:rational_extra_eq}
c_0 + c_1 m + \frac{c_2}{m+\mu} - |m| = 0.
\end{equation}
Let 
\begin{equation*}
g(\mu) := c_0(\mu) + c_1(\mu) m + \frac{c_2(\mu)}{m+\mu} - |m|,
\end{equation*}
where $c_0(\mu)$, $c_1(\mu)$, and $c_2(\mu)$ denote the values of $c_0$, $c_1$, and $c_2$, respectively, that solve the linear system $\eqref{eq:rational_3_linear_system}$ for a given value of $\mu$. That is, $g(\mu)$ is the left hand side of \eqref{eq:rational_extra_eq}, except instead of being free variables, $c_0$, $c_1$, and $c_2$ now depend on $\mu$ through the solution of \eqref{eq:rational_3_linear_system}. We solve the combined system, \eqref{eq:rational_3_linear_system} and $\eqref{eq:rational_extra_eq}$, by solving the one-dimensional nonlinear root finding problem
\begin{equation}
\label{eq:g_mu_zero}
g(\mu)=0.
\end{equation}
with bound $\mu > |m|$ and initial guess $\mu_0 = 2 |m|$. Each evaluation of $g$ implicitly requires solving the $3 \times 3$ linear system \eqref{eq:rational_3_linear_system}, and once the solution $\mu$ is found, we must solve \eqref{eq:rational_3_linear_system} one more time to get $c_0$, $c_1$, and $c_2$.
%
%Row 1 corresponds to the condition $f(M)=M$, row 2 corresponds to the condition $f(0)=0$, and row 3 corresponds to the condition $f'(0)=0$. To find the third row of $A$, we used the fact that the derivative of $f$ is given by $$f'(\lambda) = c_1 - \frac{c_2}{(\lambda+\mu)^2}.$$
%
%
%We want $f(\lambda)$ to have the following desirable properties:
%\begin{enumerate}
%	\item $f(\lambda)$ is nonnegative on $[m, M]$.
%	\item $f(\lambda) \approx \lambda$ when $\lambda$ is positive and not small.
%	\item $f(0)=0$ is the minimum of $f(\lambda)$ on $[m, M]$.
%	\item $f(\lambda) \le |\lambda|$ when $\lambda$ is negative.
%\end{enumerate}
%Property 1 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Property 2 ensures that $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. That is, the part of the spectrum of $\mathcal{A}$ that we expect to be well-approximated by the product-convolution approximation. Property 3 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ preserves the null-space (and near null-space) of $\widetilde{\mathbf{A}}_\text{sym}$. This is important because in practice, we often want to approximate operators $\mathcal{A}$ that have a large (possibly infinite) number of zero eigenvalues, and/or a sequence of eigenvalues that clusters at zero. Hessians in ill-posed inverse problems typically have both of these properties. Property 4 ensures that $f$ does not amplify erroneous negative eigenvalues $\widetilde{\mathbf{A}}_\text{sym}$. If $\mathcal{A}$ is positive semi-definite, then eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ can only be negative because of errors in the product-convolution approximation.
